---
title: "Detecting change points"
author: "Vladislav Nachev"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting change points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `cpdetectoR` package is for finding change points in sequences of responses. Putative change
points are first detected, then a statistical test at a predefined significance
level (Criterion) is applied to decide if the change point is supported or not.
Based on algorithm published by [Gallistel et al. (2004)](http://dx.doi.org/10.1073/pnas.0404965101)^[Gallistel CR, Fairhurst S, Balsam P (2004) The learning curve:
  Implications of a quantitative analysis. PNAS 101:13124-13131. doi: 10.1073/pnas.0404965101], translated from Matlab into R. 


The input can be either a vector:

```{r}
library(cpdetectoR) # load package first
set.seed(25)
cp_wrapper(c(rbinom(50, 1, 0.3), rbinom(50, 1, 0.8)), TRUE, "binomial", 2)
```

or a data frame:

```{r}
d_responses <- data.frame(Responses = c(rbinom(50, 1, 0.3), rbinom(50, 1, 0.8)))
cp_wrapper(d_responses, TRUE, "chisquare", 2)
```

For the same value of the criterion, the chi square test usually gives a higher number of change points (in this case, false positives) than the binomial test. The value of the criterion should lie between 1.3 and 6, corresponding to p values of 0.05 and 0.000001, respectively. These values are the logarithms of the odds against the null (no-change) hypothesis.

Let us look first at the included `eyeblink` data set:
```{r}
eyeblink[,] # inspect data set
```

Gallistel et al. advise against using the chi square test on these data. Indeed, with a criterion of 2, the test fails:
```{r, eval = FALSE}
cp_wrapper(eyeblink, TRUE, "chisquare", 2)
```

However, using either a higher criterion or the "binomial" test gives a result:
```{r}
cp_wrapper(eyeblink, TRUE, "chisquare", 3)

cp_wrapper(eyeblink, TRUE, "binomial", 2)
```

With ggplot we can visualize the results by first generating a `data.frame` with the cumulative responses:
```{r}
library(ggplot2) # load the ggplot package
eyeblinkdata <- data.frame(Trial = 1:length(eyeblink[,]),
                           CumRespMeasure = cumsum(eyeblink)[,])
changepoints <- cp_wrapper(eyeblink, TRUE, "binomial", 4) # save the output of the change point analysis
#generate a cumulative response vs trial plot:
ggplot(eyeblinkdata) + geom_line(aes(Trial, CumRespMeasure)) +
  geom_point(data = changepoints, aes(Trial, CumSs), size = 3)
```

Another type of plot one can look at is the average response rate per trial vs trial.
The `plusmaze` data set included with the package contains frequency data, that are again preferrably analyzed with the random rate (binomial) test.

```{r}
plusmaze[,] # inspect data set
(cp.1 <- cp_wrapper(plusmaze, TRUE, "binomial", 1.3)) #find the change points
# plot average response rate per trial
ggplot() + geom_step(data = cp.1, aes(Trial,Slopes)) +
  ylab("Average Response Rate per Trial")
# for comparison, the cumulative response vs trial plot, as in the example above:
plusmazedata <- data.frame(Trial = 1:length(plusmaze[,]),
                           CumRespMeasure = cumsum(plusmaze)[,])
ggplot(plusmazedata) + geom_line(aes(Trial, CumRespMeasure)) +
  geom_point(data = cp.1, aes(Trial, CumSs), size = 3)
```

The attached data set `hopperentry` contains hopper-entry speeds from pigeons, an example of normally distributed data. Consequently the t test can be used.

```{r}
(cp.2 <- cp_wrapper(hopperentry, TRUE, "ttest", 4)) #find the change points
# cumulative response vs trial plot
hedata <- data.frame(Trial = 1:length(hopperentry[,]),
                           CumRespMeasure = cumsum(hopperentry)[,])
pl1 <- ggplotGrob(ggplot(hedata) + geom_line(aes(Trial, CumRespMeasure)) +
  geom_point(data = cp.2, aes(Trial, CumSs), size = 3))
# plot average response rate per trial
pl2 <- ggplotGrob(ggplot(cp.2) + geom_step(aes(Trial, Slopes)) +
  ylab("Average Response Rate per Trial"))
# stack the two plots vertically using the grid package
grid::grid.draw(rbind(pl1, pl2, size = "first"))
```

An example for continuous data is the attached `matching` data set. The plots need different axes and labels:

```{r}
(cp.3 <- cp_wrapper(matching, FALSE, "binomial", 2)) #find the change points
# cumulative response vs trial plot
matchingdata <- data.frame(Events = 1:length(matching[,]),
                           Time = cumsum(matching)[,])
pl3 <- ggplotGrob(ggplot(matchingdata) + geom_line(aes(Time, Events)) +
  geom_point(data = cp.3, aes(Time, Events), size = 3))
# plot average response rate per trial
pl4 <- ggplotGrob(ggplot(cp.3) + geom_step(aes(Time, Slopes)) +
  ylab("Events per Unit Time"))
grid::grid.draw(rbind(pl3, pl4, size = "first"))
```
